---
title: "R Notebook"
output: html_notebook
---


#Once the prject is open the packages are saves in:
#  '/Users/ian.soldevilla.claramunt/Library/CloudStorage/OneDrive-KarolinskaInstitutet/ISC - Private - Karolinska Institute/5. Experiment/scRNAseq analysis/Renv_trial/renv/library/macos/R-4.5/aarch64-apple-darwin20/renv'

```{r}
install.packages("dplyr")

```
- Querying repositories for available source packages ... Done!
The following package(s) will be installed:
- cli        [3.6.6]
- dplyr      [1.2.1]
- generics   [0.1.4]
- glue       [1.8.1]
- lifecycle  [1.0.5]
- magrittr   [2.0.5]
- pillar     [1.11.1]
- pkgconfig  [2.0.3]
- R6         [2.6.1]
- rlang      [1.2.0]
- tibble     [3.3.1]
- tidyselect [1.2.1]
- utf8       [1.2.6]
- vctrs      [0.7.3]
- withr      [3.0.2]
These packages will be installed into "~/Library/CloudStorage/OneDrive-KarolinskaInstitutet/ISC - Private - Karolinska Institute/5. Experiment/scRNAseq analysis/Renv_trial/renv/library/macos/R-4.5/aarch64-apple-darwin20".


This funcituon will save in the renv.lock file all the packages and the dependencies
```{r}
renv::snapshot()
```
However dplyr is not present in this file. Bacaus the file was only installed, and the snapshot() function only stores the packages that are loaded with the function library("pakcage_name")


```{r}
#Load library in in R
library("dplyr")

#Afterwards snapshot() to save the information on the loaded libraries in the renv.lock file
#Type = all makes sure all the libraries that are loaded are saved in the renv.lock file
renv::snapshot(type = "all")
```

#by exporting the renv.lock file we can install all the packages and dependences in a new computer to recapitulate the same analysis
```{r}
#Run this exact comand to install all the packages
renv::restore(rebuild = TRUE, prompt = FALSE)

```
The packages will be downloaded and installed in the new computer
Check that this is the case buy loading library(dplyr)

```{r}
#to check if the files aare installed
renv::status()

#TO check where the libraries are downloaded
.libPaths() #SHould be inside the project directory
```

Load all ther packages with libraries("package_name") and install manually the packages with the specific verssion based on the ren.lock file
```{r}
renv::install("package_name@version_number")
```

