---
title: "Part 2 - Data processing"
Project: "BCRABL+ induced and BCRABL not induced"
output: html_notebook
---


install.packages("pheatmap")
install.packages("factoextra")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("edgeR")

install.packages("edgeR")




```{r}
## Load libraries

library(dplyr)
library(magrittr)
library(reshape2)
library(ggplot2)
library(Seurat)
library('glmGamPoi')
library(tidyverse)

```



# Data processing
```{r}
#dat.seurat_3_filtered
#dat.seurat_3


#dim(GetAssayData(dat.seurat_3, assay = "RNA", layer = "counts"))
dat.seurat_3 <- subset(dat.seurat_3, nCount_RNA > 50000)
dat.seurat_3 <- subset(dat.seurat_3, nFeature_RNA > 3000)

dat.seurat_3 <- subset(dat.seurat_3, Population %in% c("B","CD150neg","CD49bneg", "CD49bpos","GMP","LMPP","MkP","Myeloid","NK","T"))
dat.seurat_3$Population <- factor(dat.seurat_3$Population,
                                          levels = c("CD49bneg", "CD49bpos", "CD150neg", "LMPP", "GMP", "MkP", "B", "T", "NK", "Myeloid"))

# Filter genes
dat.seurat_3 <- dat.seurat_3[rowSums(dat.seurat_3@assays$RNA@layers$counts) > 300,]

dat.seurat_3 <- SCTransform(object = dat.seurat_3, verbose = FALSE, vst.flavor = "v2", return.only.var.genes = FALSE)

dat.seurat_3 <- RunPCA(dat.seurat_3, features = VariableFeatures(object = dat.seurat_3), verbose = FALSE)
dat.seurat_3 <- RunUMAP(dat.seurat_3, dims = 1:20, verbose = FALSE)

my_colours <- c("CD49bneg" = "#70193b", "CD49bpos" = "#ff6484", "CD150neg" = "#e09e47", "LMPP" = "#ff7206", "GMP" = "#70bd5e", "MkP" = "#9620d7", "B" = "#014784", "T" = "#9ca6f1", "NK" = "#00bbd6", "Myeloid" = "#4a6800", "transplanted" = "grey")

DimPlot(dat.seurat_3, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Population", cols = my_colours)
DimPlot(dat.seurat_3, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Group")
```

```{r}
p <- DimPlot(dat.seurat_3, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Population", cols = my_colours)

HoverLocator(p, information = FetchData(object = dat.seurat_3, vars = 'Well'))
```

# Make clusters and find markers per hasshed population
```{r}
dat.seurat_3.markers.population <- FindAllMarkers(dat.seurat_3, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5, group.by = "Population")
dat.seurat_3.markers.population <- dat.seurat_3.markers.population[dat.seurat_3.markers.population$p_val_adj < 0.01, ]
top10_3.population <- dat.seurat_3.markers.population %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

top10_3.population$gene_name <- gene_names$gene_name[match(top10_3.population$gene,gene_names$gene_id,top10_3.population$gene)]

DoHeatmap(dat.seurat_3, features = top10_3.population$gene, group.by = "Population") +
  scale_fill_gradientn(colors = c("blue", "black", "yellow")) +
  theme(text = element_text(size = 10)) +
  scale_y_discrete(labels = rev(c(gene_names$gene_name[match(top10_3.population$gene,gene_names$gene_id,top10_3.population$gene)])))
```

# Make clusters and find markers per group
```{r}
dat.seurat_3.markers.group <- FindAllMarkers(dat.seurat_3, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5, group.by = "Group")
dat.seurat_3.markers.group <- dat.seurat_3.markers.group[dat.seurat_3.markers.group$p_val_adj < 0.01, ]
top10_3.group <- dat.seurat_3.markers.group %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

top10_3.group$gene_name <- gene_names$gene_name[match(top10_3.group$gene,gene_names$gene_id,top10_3.group$gene)]

DoHeatmap(dat.seurat_3, features = top10_3.group$gene, group.by = "Group") +
  scale_fill_gradientn(colors = c("blue", "black", "yellow")) +
  theme(text = element_text(size = 10)) +
  scale_y_discrete(labels = rev(c(gene_names$gene_name[match(top10_3.group$gene,gene_names$gene_id,top10_3.group$gene)])))
```
#Pseudobulk reduction

#Run Pearson correlation analysis between groups

```{r}
#Check bulk RNA seq pipelie for PEarson correlation analysis and scRNAseq script fot pseudobulk
#From chat gpt

#Extract expression data

expr <- GetAssayData(
  dat.seurat_3,
  assay = "RNA",
  layer = "counts"   # log-normalized expression
)

#Aggregate expression by group (pseudobulk)

dat.seurat_3@meta.data

pseudo_3 <- AggregateExpression(dat.seurat_3, assays = "RNA", return.seurat = T, group.by = "Group")

expr <- GetAssayData(
  pseudo_3,
  assay = "RNA",
  layer = "counts"   # log-normalized expression
)
Head(expr)

pseudo_neurons$celltype.stim <- paste(pseudo_neurons$celltypes, pseudo_neurons$status, sep = "_")

Idents(pseudo_neurons) <- "celltype.stim"

pseudo_3 <- FindMarkers(object = pseudo_3, 
                         ident.1 = "BCR-ABL-7d", 
                         ident.2 = "BCR-WT",
                         ident.2 = "WT",
                         test.use = "DESeq2")


FindMarkers(
  object = pseudo_3,
  ident.1 = "BCR-ABL-7d",
  ident.2 = "BCR-WT",
  test.use = "MAST"
)




avg.expr <- AverageExpression(
  dat.seurat_3,
  assays = "RNA",
  layer = "data",
  group.by = group.var,
  verbose = FALSE
)$RNA


#Filter genes before correlation

genes.use <- VariableFeatures(dat.seurat_3)
avg.expr <- avg.expr[genes.use, ]

#Run Pearson correlation analysis

cor.matrix <- cor(
  avg.expr,
  method = "pearson"
)

#Visualize correlation

#Option 1
heatmap(
  cor.matrix,
  symm = TRUE,
  margins = c(8, 8)
)


#Option 2
library(pheatmap)
pheatmap(
  cor.matrix,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  clustering_method = "average"
)



cor(
  AverageExpression(dat.seurat_3, group.by = "Group")$RNA[
    VariableFeatures(dat.seurat_3),
  ],
  method = "pearson"
)




```

#Pearson correlation coeficient

```{r, fig.width=8, fig.height=6}
library(pheatmap)
as.matrix(pseudo_3$RNA)


pseudo_3 <- AggregateExpression(dat.seurat_3, assays = "RNA", group.by = "Group")
pseudo_3_group_population <- AggregateExpression(dat.seurat_3, assays = "RNA", return.seurat = T, group.by = c("Group","Population"))

pseudo_3_group_population$group.population <- paste(pseudo_3_group_population$Group, pseudo_3_group_population$Population, sep = "_")

pseudo_3_group_population@meta.data

Idents(pseudo_3_group_population) <- "group.population"

pseudo_3_group_population_matrix <- as.matrix(GetAssayData(pseudo_3_group_population, assay = "RNA", layer = "counts"))

cor_matrix_pearson <- cor(pseudo_3_group_population_matrix, method = "pearson")
cor_matrix_pearson
pheatmap(cor_matrix_pearson, main = "Pearson Correlation of Samples", cluster_rows = FALSE, cluster_cols = FALSE)


```

#PCA of pseudobulk agrupation

```{r}
#demultiplex data

#plot PCA grouped with hashed samples and group

library(factoextra)
library(edgeR)



#Not working
mydgelist <- DGEList(pseudo_3_group_population_matrix, ,group=c(3,3,3,3,3,3,3,3,3,3,2,2,2,2,2,2,2,2,2,2,1,1,1,1,1,1))

mydgelist <- DGEList(pseudo_3_group_population_matrix)

#mydgelist_samples <- mydgelist$samples %>% rownames_to_column("sample") %>% separate(sample, into =c("genotype", "replicate"))

mydgelist_samples <- mydgelist$samples %>%  rownames_to_column("sample") %>%
  separate(sample, into = c("Group", "Population"), sep = "_")

mydgelist_cpm <- cpm(counttablematrix_filt)

#Coloured by Group

res.pca <- prcomp(t(mydgelist$counts))
#colors <- as.factor(str_split(rownames(mydgelist$samples), "_", simplify = TRUE)[,1])
colors <- mydgelist_samples$Group

fviz_pca_ind(res.pca,
             palette = "Set1",
             col.ind = mydgelist_samples$Group,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )


#Cd49b+, Cd49b-, Cd150-, MkP, LMPP, GMP

#Coloured by Population
colors <- mydgelist_samples$Population

fviz_pca_ind(res.pca,
             palette = "Set1",
             col.ind = mydgelist_samples$Population,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )

#Coloured by Group and Population

fviz_pca_ind(res.pca,
             palette = "Set1",
             col.ind = mydgelist_samples$Group,
             habillage = mydgelist_samples$Population,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )  +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9))


#Lin; B-cells, T-cells, M-cells, NK-cells





fviz_pca_ind(
  res.pca,
  palette = "Set1",
  repel = TRUE,
  title = "PCA plot of human data"
)

fviz_pca_ind(
  res.pca,
  habillage = pseudo_3_group_population$group,  # <- factor variable
  palette = "Set1",
  repel = TRUE,
  title = "PCA plot of human data"
)

```

