---
title: "Part 4 - Data processing combined batches"
output: html_notebook
---


#When opening a project that was created in another computer we need to run the following comand in o
```{r}
renv::restore()
```


```{r}
## Load libraries

library(dplyr)
library(magrittr)
library(reshape2)
library(ggplot2)
library(Seurat)
library('glmGamPoi')
library(tidyverse)

library(PCAtools)
library(pheatmap)
library(factoextra)
library(edgeR)
library(scales)
library(writexl)


```

#Load pilot plate
```{r}

df_1 <- read.table("../Data/RawData/SS3xpress_B-ABL_pilot_20240115/zUMIs_output/Xpress.readcounts.intron_exon.txt", header = TRUE, sep = "\t")

barcode_annotation_1 <-read.table("../Data/RawData/SS3xpress_B-ABL_pilot_20240115/zUMIs_output/barcode_annotation.txt", header = T, sep = "\t")

gene_names_1 <- read.table("../Data/RawData/SS3xpress_B-ABL_pilot_20240115/zUMIs_output/Xpress.gene_names.txt", header = TRUE, sep = "\t")

barcode_annotation_1$Well_plateID <- paste(barcode_annotation_1$WellID, barcode_annotation_1$plateBC, sep = "_")
colnames(df_1) <- barcode_annotation_1$Well_plateID[match(colnames(df_1),barcode_annotation_1$XC_DNBPE)]
df_1 <- df_1[,order(colnames(df_1))]


## Metadata pilot plate
metadata_1 <- read.csv2("../Data/RawData/SS3xpress_B-ABL_pilot_20240115/SS3xpress_metadata.csv", header = TRUE)
metadata_1 <- metadata_1[order(metadata_1$Plate.ID),]
metadata_1 <- metadata_1[order(metadata_1$Well),]
rownames(metadata_1) <- colnames(df_1)



```

# Load data batch 3 - BCR-ABL 051-052
```{r}
## Batch 3
df_3 <- read.table("../Data/RawData/SS3xpress_B-ABL_051-052_20250617/zUMIs_output/Xpress.readcounts.intron_exon.txt", header = TRUE, sep = "\t")

barcode_annotation_3 <-read.table("../Data/RawData/SS3xpress_B-ABL_051-052_20250617/zUMIs_output/barcode_annotation.txt", header = T, sep = "\t")
gene_names_3 <- read.table("../Data/RawData/SS3xpress_B-ABL_051-052_20250617/zUMIs_output/Xpress.gene_names.txt", header = TRUE, sep = "\t")

barcode_annotation_3$Well_plateID <- paste(barcode_annotation_3$WellID, barcode_annotation_3$plateBC, sep = "_")
colnames(df_3) <- barcode_annotation_3$Well_plateID[match(colnames(df_3),barcode_annotation_3$XC_DNBPE_TS)]
df_3 <- df_3[,order(colnames(df_3))]

## Metadata batch 3
metadata_3 <- read.csv2("../Data/RawData/SS3xpress_B-ABL_051-052_20250617/SS3xpress_metadata.csv", header = TRUE)
metadata_3 <- metadata_3[order(metadata_3$Plate.ID),]
metadata_3 <- metadata_3[order(metadata_3$Well),]
rownames(metadata_3) <- colnames(df_3)

# Check if Gene names are the same and if yes only keep one file
#all.equal(gene_names_1,gene_names_2)
all.equal(gene_names_1,gene_names_3)
gene_names <- gene_names_3
rm(gene_names_1, gene_names_3)


```

#Asign mice ID in metadata

```{r}


#We will add in the metadaa the nummber of the batch
metadata_1$batch <- 1
metadata_3$batch <- 3

#We will merge the metadata for every sample
metadata_combined <- rbind(metadata_1, metadata_3)

#We will annotate the ID of the mice in the metadata based on the sorting notes and the plate information

metadata_combined <- metadata_combined %>%
  mutate(Mouse.ID = case_when(
    Plate.ID == 11280577 & Column %in% c(1:7, 22) ~ "Mouse P23-3-1-2 WT",
    Plate.ID == 11280577 & Column %in% c(8:14, 23) ~ "Mouse P23-3-1-3 WT",
    Plate.ID == 11280577 & Column %in% c(15:21, 24) ~ "Mouse P23-4-2-4 WT",

    Plate.ID == 11280585 & Column %in% 1:6 ~ "Mouse P23-5-1-8 WT",
    Plate.ID == 11280585 & Column %in% c(19,20,21) & Row %in% 1:8 ~ "Mouse P23-5-1-8 WT",

    Plate.ID == 11280585 & Column %in% 7:12 ~ "Mouse P23-5-1-9 WT",
    Plate.ID == 11280585 & Column %in% c(19,20,21) & Row %in% 9:16 ~ "Mouse P23-5-1-9 WT",

    Plate.ID == 11280585 & Column %in% c(13,14,15) ~ "Mouse P23-5-1-10 WT",
    Plate.ID == 11280585 & Column %in% c(22,23,24) & Row %in% 1:8 ~ "Mouse P23-5-1-10 WT",

    Plate.ID == 11280585 & Column %in% c(16,17,18) ~ "Mouse P23-5-1-11 WT",
    Plate.ID == 11280585 & Column %in% c(22,23,24) & Row %in% 9:16 ~ "Mouse P23-5-1-11 WT",
    
    Plate.ID == 348  & Group == "BCR_ABL_7d" ~ "Mouse 1 BCR-ABL-7d",
    Plate.ID == 348  & Group == "WT" ~ "Mouse 2 WT",
    
    Plate.ID == 11280344  & Group == "BCR_ABL_7d" ~ "Mouse 23HQ25_8_M BCR_ABL_7d",
    Plate.ID == 11280344  & Group == "BCR_WT" ~ "Mouse 23HQ25_7_F BCR_WT",
    
    Plate.ID == 11280345  & Group == "BCR_ABL_7d" ~ "Mouse 23HQ25_8_M BCR_ABL_7d",
    Plate.ID == 11280345  & Group == "BCR_WT" ~ "Mouse 23HQ25_7_F BCR_WT",

    TRUE ~ NA_character_
  ))
```


## Load data into Seurat 
```{r}

#First we will create a seurat object containing independent samples
dat.seurat_1 <- CreateSeuratObject(counts = df_1, meta.data = metadata_1, min.cells = 10, min.features = 0)

dat.seurat_3 <- CreateSeuratObject(counts = df_3, meta.data = metadata_3, min.cells = 10, min.features = 0)

#In order to create independent samples first we need to merge the count matrixes (df_1 and df_2)

df_combined <- merge(df_1, df_3, by = "row.names", all = FALSE)
rownames(df_combined) <- df_combined$Row.names
df_combined <- df_combined[,-1]


#Finally we will create an seurat object containing the from the combined count matrix and the combined metadata

dat.seurat.combined <- CreateSeuratObject(counts = df_combined, meta.data = metadata_combined, min.cells = 10, min.features = 0)


#Remove the plate "11280345" from the batch 3, it is the only plate that sorted B-cells, T-cells, M-cells, NK-cells, therefore I will focuss in the Cd49b+, Cd49b-, Cd150-, MkP, LMPP and GMP populations
 
dat.seurat.combined <- subset(dat.seurat.combined, Plate.ID != "11280345")


#rm(df_1, df_2, df_3, metadata_1, metadata_2, metadata_3, df_combined, metadata_combined)

unique(dat.seurat.combined@meta.data$Plate.ID)

#In order to easy the comprehenssion I will change the name of the plates: 

dat.seurat.combined@meta.data <- dat.seurat.combined@meta.data %>%
  mutate(Plate.ID = recode(Plate.ID,
    "348" = "Plate.1",
    "11280344" = "Plate.2",
    "11280577" = "Plate.4",
    "11280585" = "Plate.5"
  ))

dat.seurat.combined@meta.data <- dat.seurat.combined@meta.data %>%
  mutate(Plate.Group = paste(Plate.ID, Group, sep = "-"))

#Count mitochondrial percentage
mt_genes <- gene_names$gene_id[grep("^mt-", gene_names$gene_name)]
dat.seurat.combined@meta.data$percent.mt <- PercentageFeatureSet(dat.seurat.combined, features = mt_genes)

dat.seurat.combined@meta.data
#write_xlsx(dat.seurat.combined@meta.data, "../outputs/my_dataframe.xlsx")

```


```{r}
x <- dat.seurat.combined@meta.data
```


```{r}
# Visualize QC metrics as a violin plot
VlnPlot(dat.seurat.combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3) +
    scale_y_log10()

VlnPlot(dat.seurat.combined, features = "nCount_RNA") + scale_y_log10() + 
  geom_hline(yintercept = 50000, linetype = "dashed", color = "red")+ ggtitle("Ttotal RNA count per cell")

VlnPlot(dat.seurat.combined, features = "nFeature_RNA") + scale_y_log10() + 
  geom_hline(yintercept = 3000, linetype = "dashed", color = "red")+ ggtitle("Number of genes per cell")

VlnPlot(dat.seurat.combined, features = "percent.mt") + 
  geom_hline(yintercept = 10, linetype = "dashed", color = "red")+ ggtitle(" Mitochondrial gene counts percentage")

#dim(GetAssayData(dat.seurat.combined, assay = "RNA", layer = "counts"))
dat.seurat.combined <- subset(dat.seurat.combined, nCount_RNA > 50000)
dat.seurat.combined <- subset(dat.seurat.combined, nFeature_RNA > 3000)
dat.seurat.combined <- subset(dat.seurat.combined, percent.mt <10)

VlnPlot(dat.seurat.combined, features = "nCount_RNA") + scale_y_log10() + ggtitle("Ttotal RNA count per cell - Filtered")
VlnPlot(dat.seurat.combined, features = "nFeature_RNA") + scale_y_log10() + ggtitle("Number of genes per cell - Filtered")
VlnPlot(dat.seurat.combined, features = "percent.mt") + ggtitle(" Mitochondrial gene counts percentage - Filtered")


```


#Data processing
```{r}

#write which population to choose
dat.seurat.combined <- subset(dat.seurat.combined, Population %in% c("B","CD150neg","CD49bneg", "CD49bpos","GMP","LMPP","MkP","Myeloid","NK","T"))
dat.seurat.combined$Population <- factor(dat.seurat.combined$Population,
                                          levels = c("CD49bneg", "CD49bpos", "CD150neg", "LMPP", "GMP", "MkP", "B", "T", "NK", "Myeloid"))

# Filter genes - this functions will filter out genes that are not expressed among all single cells across all samples with al least 300 counts
#May be this funtion is not ver 
#dat.seurat.combined <- dat.seurat.combined[rowSums(dat.seurat.combined@assays$RNA@layers$counts) > 300,]

# Split object for later integration
dat.seurat.combined[["RNA"]] <- split(dat.seurat.combined[["RNA"]], f = dat.seurat.combined$Mouse.ID)

# SC Transform
dat.seurat.combined <- SCTransform(object = dat.seurat.combined, verbose = FALSE, vst.flavor = "v2", return.only.var.genes = FALSE)

# Dimensionality reduction
dat.seurat.combined <- RunPCA(dat.seurat.combined, features = VariableFeatures(object = dat.seurat.combined), verbose = FALSE)
dat.seurat.combined <- RunUMAP(dat.seurat.combined, dims = 1:20, verbose = FALSE)

my_colours <- c("CD49bneg" = "#70193b", "CD49bpos" = "#ff6484", "CD150neg" = "#e09e47", "LMPP" = "#ff7206", "GMP" = "#70bd5e", "MkP" = "#9620d7")

DimPlot(dat.seurat.combined, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Population", cols = my_colours)
DimPlot(dat.seurat.combined, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Group")
DimPlot(dat.seurat.combined, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Plate.ID")
DimPlot(dat.seurat.combined, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Plate.Group")
DimPlot(dat.seurat.combined, reduction = "umap", label = FALSE, repel = TRUE, group.by = "Mouse.ID")

```

# Integrate plates based on Batch with CCA or harmony Integration
```{r}

dat.seurat.combined <- IntegrateLayers(dat.seurat.combined, method = CCAIntegration, normalization.method = "SCT", orig.reduction = "pca", new.reduction = "cca", verbose = FALSE)
dat.seurat.combined <- IntegrateLayers(dat.seurat.combined, method = HarmonyIntegration, normalization.method = "SCT", new.reduction = "harmony", verbose = FALSE)

# re-join layers after integration, when we want ot have all the assays in the same matrix
#dat.seurat.combined[["RNA"]] <- JoinLayers(dat.seurat.combined[["RNA"]])

dat.seurat.combined <- RunUMAP(dat.seurat.combined, reduction = "cca", dims = 1:20, verbose = FALSE, reduction.name = "umap_cca")
dat.seurat.combined <- RunUMAP(dat.seurat.combined, reduction = "harmony", dims = 1:20, verbose = FALSE, reduction.name = "umap_harmony")

#PLot UMAPs with cca integration
DimPlot(dat.seurat.combined, reduction = "umap_cca", label = FALSE, repel = TRUE, group.by = "Population", cols = my_colours)
DimPlot(dat.seurat.combined, reduction = "umap_cca", label = FALSE, repel = TRUE, group.by = "Group")
DimPlot(dat.seurat.combined, reduction = "umap_cca", label = FALSE, repel = TRUE, group.by = "Plate.ID")
DimPlot(dat.seurat.combined, reduction = "umap_cca", label = FALSE, repel = TRUE, group.by = "Plate.Group")
DimPlot(dat.seurat.combined, reduction = "umap_cca", label = FALSE, repel = TRUE, group.by = "Mouse.ID")

#Plot UMAPs with Harmony integration
DimPlot(dat.seurat.combined, reduction = "umap_harmony", label = FALSE, repel = TRUE, group.by = "Population", cols = my_colours)
DimPlot(dat.seurat.combined, reduction = "umap_harmony", label = FALSE, repel = TRUE, group.by = "Group")
DimPlot(dat.seurat.combined, reduction = "umap_harmony", label = FALSE, repel = TRUE, group.by = "Plate.ID")
DimPlot(dat.seurat.combined, reduction = "umap_harmony", label = FALSE, repel = TRUE, group.by = "Plate.Group")
DimPlot(dat.seurat.combined, reduction = "umap_harmony", label = FALSE, repel = TRUE, group.by = "Mouse.ID")

```

```{r}
dat.seurat.combined <- FindNeighbors(dat.seurat.combined, reduction = "cca", dims = 1:20, verbose = FALSE)
dat.seurat.combined <- FindClusters(dat.seurat.combined, resolution = 0.6, verbose = FALSE, cluster.name = "cca_clusters")
```

# Make clusters and find markers per hasshed population
```{r}
dat.seurat.combined.markers.population <- FindAllMarkers(dat.seurat.combined, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5, group.by = "Population")
dat.seurat.combined.markers.population <- dat.seurat.combined.markers.population[dat.seurat.combined.markers.population$p_val_adj < 0.01, ]
top10_combined.population <- dat.seurat.combined.markers.population %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

top10_combined.population$gene_name <- gene_names$gene_name[match(top10_combined.population$gene,gene_names$gene_id,top10_combined.population$gene)]

DoHeatmap(dat.seurat.combined, features = top10_combined.population$gene, group.by = "Population") +
  scale_fill_gradientn(colors = c("blue", "black", "yellow")) +
  theme(text = element_text(size = 10)) +
  scale_y_discrete(labels = rev(c(gene_names$gene_name[match(top10_combined.population$gene,gene_names$gene_id,top10_combined.population$gene)])))
```

# Make clusters and find markers per group
```{r}
dat.seurat.combined.markers.group <- FindAllMarkers(dat.seurat.combined, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5, group.by = "Group")
dat.seurat.combined.markers.group <- dat.seurat.combined.markers.group[dat.seurat.combined.markers.group$p_val_adj < 0.01, ]
top10_combined.group <- dat.seurat.combined.markers.group %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

top10_combined.group$gene_name <- gene_names$gene_name[match(top10_combined.group$gene,gene_names$gene_id,top10_combined.group$gene)]

DoHeatmap(dat.seurat.combined, features = top10_combined.group$gene, group.by = "Group") +
  scale_fill_gradientn(colors = c("blue", "black", "yellow")) +
  theme(text = element_text(size = 10)) +
  scale_y_discrete(labels = rev(c(gene_names$gene_name[match(top10_combined.group$gene,gene_names$gene_id,top10_combined.group$gene)])))
```

#Pearson correlation coeficient

```{r, fig.width=8, fig.height=6}

library(pheatmap)

#Aggregate samples using pseudobulk reduction - the sample will be grouped by "Group"
seurat_pseudo_combined_group <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = TRUE,  group.by = "Group")
seurat_pseudo_combined_group@meta.data

#Get the count matrix without populations
seurat_pseudo_combined_group_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_group, assay = "RNA", layer = "counts"))

#Calculate Pearson corelation coefficient (p) without populations
cor_matrix_pearson_group <- cor(seurat_pseudo_combined_group_counts, method = "pearson")
cor_matrix_pearson_group

#Generate heatmap with p
pheatmap(cor_matrix_pearson_group, main = "Pearson Correlation Analysis between Groups", cluster_rows = FALSE, cluster_cols = FALSE)


```

#Pearson's correlation coeficient of Group, Population and PLate
```{r}
#Aggregate samples using pseudobulk reduction - Populations will be grouped by the genotype and hashed markersand "Group" and Populations"
#AggregateExpression() created a new orig.ident column with the combination of the gorups set in group.by()

seurat_pseudo_combined_group_population <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = T, group.by = c("Group","Population"))

#seurat_pseudo_combined_group_population$group.population <- paste(seurat_pseudo_combined_group_population$Group, pseudo_combined_group_population$Population, sep = "_")
#Idents(pseudo_combined_group_population) <- "group.population"


#Get the count matrix
pseudo_combined_group_population_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_group_population, assay = "RNA", layer = "counts"))

#Perform Pearson corelation analysis of al groups and populations
cor_matrix_pearson_group_population <- cor(pseudo_combined_group_population_counts, method = "pearson")
cor_matrix_pearson_group_population
pheatmap(cor_matrix_pearson_group_population, main = "Pearson Correlation Analysis of Populations", cluster_rows = FALSE, cluster_cols = FALSE)

#PCA
mydgelist <- DGEList(pseudo_combined_group_population_counts)

#First add the row names into a column
mydgelist$samples <- mydgelist$samples %>%  rownames_to_column("sample") 

#Duplicate this column
mydgelist$samples$Group.Population <- mydgelist$samples$sample

#Separate this column into Group and Population and bring the initial rowname back to the rowname of the metadata 
mydgelist$samples <- mydgelist$samples %>% separate(Group.Population, into = c("Group","Population"), sep = "_") %>% column_to_rownames("sample")
#Run principal component analysis
res.pca <- pca(mydgelist$counts, metadata = mydgelist$samples, removeVar = 0.1) #To run this funtion the metadata object need to have the name of the samples as a rownames()
#Generate PCA plot colouring by Population and the shape by Group
biplot(res.pca, 
       shape = "Population", # to asing wich value in the metadata will be used to group the the shape
       colby = "Group", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

```

#Pearson's coerelation coeficient of Plate and Group

```{r}
#Aggregate samples using pseudobulk reduction - Populations will be grouped by the genotype and hashed markersand "Group" and Populations"
#AggregateExpression() created a new orig.ident column with the combination of the gorups set in group.by()

seurat_pseudo_combined_group_plate <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = T, group.by = c( "Plate.ID","Group"))

#Get the count matrix
seurat_pseudo_combined_group_plate_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_group_plate, assay = "RNA", layer = "counts"))

#Perform Pearson corelation analysis of al groups and populations
cor_matrix_pearson_group_plate <- cor(seurat_pseudo_combined_group_plate_counts, method = "pearson")
cor_matrix_pearson_group_plate
pheatmap(cor_matrix_pearson_group_plate, main = "Pearson Correlation Analysis of Populations", cluster_rows = FALSE, cluster_cols = FALSE)

#PCA
mydgelist <- DGEList(seurat_pseudo_combined_group_plate_counts)

#First add the row names into a column
mydgelist$samples <- mydgelist$samples %>%  rownames_to_column("sample") 

#Duplicate this column
mydgelist$samples$Group.Population <- mydgelist$samples$sample

#Separate this column into Group and Population and bring the initial rowname back to the rowname of the metadata 
mydgelist$samples <- mydgelist$samples %>% separate(Group.Population, into = c("Plate.ID","Group"), sep = "_") %>% column_to_rownames("sample")
#Run principal component analysis
res.pca <- pca(mydgelist$counts, metadata = mydgelist$samples, removeVar = 0.1) #To run this funtion the metadata object need to have the name of the samples as a rownames()
#Generate PCA plot colouring by Population and the shape by Group
biplot(res.pca, 
       shape = "Plate.ID", # to asing wich value in the metadata will be used to group the the shape
       colby = "Group", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)
```

```{r, fig.hight=10, fig-width=12}


#Aggregate samples using pseudobulk reduction - Populations will be grouped by the genotype and hashed markersand "Group" and Populations"
#AggregateExpression() created a new orig.ident column with the combination of the gorups set in group.by()

seurat_pseudo_combined_group_population_plate <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = T, group.by = c("Group","Population", "Plate.ID"))

#seurat_pseudo_combined_group_population_plate$group.population <- paste(seurat_pseudo_combined_group_population_plate$Group, pseudo_combined_group_population_plate$Population, sep = "_")
#Idents(pseudo_combined_group_population_plate) <- "group.population"


#Get the count matrix
pseudo_combined_group_population_plate_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_group_population_plate, assay = "RNA", layer = "counts"))

#Perform Pearson corelation analysis of al groups and populations
cor_matrix_pearson_group_population_plate <- cor(pseudo_combined_group_population_plate_counts, method = "pearson")
cor_matrix_pearson_group_population_plate
pheatmap(cor_matrix_pearson_group_population_plate, main = "Pearson Correlation Analysis of Populations", cluster_rows = FALSE, cluster_cols = FALSE)

#PCA
mydgelist <- DGEList(pseudo_combined_group_population_plate_counts)

#First add the row names into a column
mydgelist$samples <- mydgelist$samples %>%  rownames_to_column("sample") 

#Duplicate this column
mydgelist$samples$Group.Population <- mydgelist$samples$sample

#Separate this column into Group and Population and bring the initial rowname back to the rowname of the metadata 
mydgelist$samples <- mydgelist$samples %>% separate(Group.Population, into = c("Group","Population", "Plate.ID"), sep = "_") %>% column_to_rownames("sample")
#Run principal component analysis
res.pca <- pca(mydgelist$counts, metadata = mydgelist$samples, removeVar = 0.1) #To run this funtion the metadata object need to have the name of the samples as a rownames()

#Generate PCA plot colouring by Population and the shape by Group
biplot(res.pca, 
       shape = "Plate.ID", # to asing wich value in the metadata will be used to group the the shape
       colby = "Group", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

biplot(res.pca, 
       shape = "Plate.ID", # to asing wich value in the metadata will be used to group the the shape
       colby = "Population", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

biplot(res.pca, 
       shape = "Population", # to asing wich value in the metadata will be used to group the the shape
       colby = "Group", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)
```



#Pearson's correlation coeficient Mouse.ID, Group and Plate ID
```{r, fig.hight=10, fig-width=12}
#Aggregate samples using pseudobulk reduction - Populations will be grouped by the genotype and hashed markersand "Group" and Populations"
#AggregateExpression() created a new orig.ident column with the combination of the gorups set in group.by()

seurat_pseudo_combined_Mouse.ID <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = T, group.by = c("Mouse.ID"))

#Get the count matrix
pseudo_combined_Mouse.ID_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_Mouse.ID, assay = "RNA", layer = "counts"))

#Perform Pearson corelation analysis of al groups and populations
cor_matrix_pearson_Mouse.ID <- cor(pseudo_combined_Mouse.ID_counts, method = "pearson")
cor_matrix_pearson_Mouse.ID[upper.tri(cor_matrix_pearson_Mouse.ID)] <- NA
pheatmap(cor_matrix_pearson_Mouse.ID, main = "Pearson Correlation Analysis of Populations", cluster_rows = FALSE, cluster_cols = FALSE)
cor_matrix_pearson_Mouse.ID


#Aggregrate the seurat data by Mouse.ID, Group, and Plate.ID
seurat_pseudo_combined_Mouse.ID <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = T, group.by = c("Mouse.ID", "Group", "Plate.ID"))

#Get the count matrix
pseudo_combined_Mouse.ID_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_Mouse.ID, assay = "RNA", layer = "counts"))
#PCA
mydgelist <- DGEList(pseudo_combined_Mouse.ID_counts)

#First add the row names into a column
mydgelist$samples <- mydgelist$samples %>% rownames_to_column("sample") 

#Duplicate this column
mydgelist$samples$Sample <- mydgelist$samples$sample

#Separate this column into Group and Population and bring the initial rowname back to the rowname of the metadata 
mydgelist$samples <- mydgelist$samples %>% separate(Sample, into = c("Mouse.ID", "Group", "Plate.ID"), sep = "_") %>% column_to_rownames("sample")
#Run principal component analysis
res.pca <- pca(mydgelist$counts, metadata = mydgelist$samples, removeVar = 0.1) #To run this funtion the metadata object need to have the name of the samples as a rownames()


#Generate PCA plot colouring by Population and the shape by Group
biplot(res.pca,
       shape = "Group",   # to asing wich value in the metadata will be used to group the the shape
       colby = "Mouse.ID", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Group - Mouse.ID") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

biplot(res.pca,
       shape = "Plate.ID",   # to asing wich value in the metadata will be used to group the the shape
       colby = "Mouse.ID", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Mouse.ID - Plate.ID") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

```


```{r}


#Aggregate samples using pseudobulk reduction - Populations will be grouped by the genotype and hashed markersand "Group" and Populations"
#AggregateExpression() created a new orig.ident column with the combination of the gorups set in group.by()

seurat_pseudo_combined_group_population_plate_mouse <- AggregateExpression(dat.seurat.combined, assays = "RNA", return.seurat = T, group.by = c("Population", "Mouse.ID", "Group"))

#seurat_pseudo_combined_group_population_plate$group.population <- paste(seurat_pseudo_combined_group_population_plate$Group, pseudo_combined_group_population_plate$Population, sep = "_")
#Idents(pseudo_combined_group_population_plate) <- "group.population"


#Get the count matrix
pseudo_combined_group_population_plate_mouse_counts <- as.matrix(GetAssayData(seurat_pseudo_combined_group_population_plate_mouse, assay = "RNA", layer = "counts"))

#Perform Pearson corelation analysis of al groups and populations
cor_matrix_pearson_group_population_plate_mouse <- cor(pseudo_combined_group_population_plate_mouse_counts, method = "pearson")
cor_matrix_pearson_group_population_plate_mouse
pheatmap(cor_matrix_pearson_group_population_plate_mouse, main = "Pearson Correlation Analysis of Populations", cluster_rows = FALSE, cluster_cols = FALSE)

#PCA
mydgelist <- DGEList(pseudo_combined_group_population_plate_mouse_counts)

#First add the row names into a column
mydgelist$samples <- mydgelist$samples %>%  rownames_to_column("sample") 

#Duplicate this column
mydgelist$samples$Group.Population <- mydgelist$samples$sample

#Separate this column into Group and Population and bring the initial rowname back to the rowname of the metadata 
mydgelist$samples <- mydgelist$samples %>% separate(Group.Population, into = c("Population", "Mouse.ID", "Group"), sep = "_") %>% column_to_rownames("sample")
#Run principal component analysis
res.pca <- pca(mydgelist$counts, metadata = mydgelist$samples, removeVar = 0.1) #To run this funtion the metadata object need to have the name of the samples as a rownames()
#Generate PCA plot colouring by Population and the shape by Group
biplot(res.pca, 
       shape = "Population", # to asing wich value in the metadata will be used to group the the shape
       colby = "Mouse.ID", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

biplot(res.pca, # to asing wich value in the metadata will be used to group the the shape
       colby = "Mouse.ID", #to asign which value in the metadata will be used to grop the color
       shape = "Group",
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)

```



#PCA with PCAtools (Bioconductor)


  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')

  BiocManager::install('PCAtools')



```{r}

#Create a DGE object to store the count matrix
mydgelist <- DGEList(pseudo_combined_group_population_plate_mouse_counts)

#See what is stored in this object
names(mydgelist)

#In order to run PCA we first need to annotate the Population and Group in separate columns. The row names has anotated the POpulation and the Group in the same value separated by a "_"
rownames(mydgelist$samples)

#First add the row names into a column
mydgelist$samples <- mydgelist$samples %>%  rownames_to_column("sample") 

#Duplicate this column
mydgelist$samples$Group.Population <- mydgelist$samples$sample

#Separate this column into Group and Population and bring the initial rowname back to the rowname of the metadata 
mydgelist$samples <- mydgelist$samples %>% separate(Group.Population, into = c("Group", "Mouse.ID"), sep = "_") %>% column_to_rownames("sample")


#Run principal component analysis

res.pca <- pca(mydgelist$counts, metadata = mydgelist$samples, removeVar = 0.1) #To run this funtion the metadata object need to have the name of the samples as a rownames()
res.pca

#Generate PCA plot colouring by Population and the shape by Group

biplot(res.pca, 
       shape = "Group", # to asing wich value in the metadata will be used to group the the shape
       colby = "Population", #to asign which value in the metadata will be used to grop the color
       lab = NULL,                        #Remove the label of the points
       legendPosition = "right",          #indicate where the legend is goinf to be, whwn activated it automatically will generate legends for the color and shape
       pointSize = 2, 
       title = "PCA Groups and Populations") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)
  #to write the labs in the x axis in scientific numbering

#biplot(res.pca, colby = "Group")
biplot(res.pca, 
       colby = "Group", 
       lab = NULL,                        
       legendPosition = "right",          
       pointSize = 2, 
       title = "PCA Groups") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)


biplot(res.pca, 
       shape = "Plate.ID",
       colby = "Population",
       lab = NULL,         
       legendPosition = "right",
       pointSize = 2,
       title = "PCA Plate.ID and Populations") +
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)


#biplot(res.pca, colby = "Group")
biplot(res.pca, 
       colby = "Mouse.ID", 
       lab = NULL,                        
       legendPosition = "right",          
       pointSize = 2, 
       title = "PCA Groups") + 
  scale_x_continuous(labels = scientific)  +
  scale_y_continuous(labels = scientific)


```


#PCA of pseudobulk agrupation

```{r}
#demultiplex data

#plot PCA grouped with hashed samples and group





#Not working
#mydgelist <- DGEList(pseudo_combined_group_population_matrix, ,group=c(3,3,3,3,3,3,3,3,3,3,2,2,2,2,2,2,2,2,2,2,1,1,1,1,1,1))

mydgelist <- DGEList(pseudo_combined_group_population_counts)
mydgelist$samples

#mydgelist_samples <- mydgelist$samples %>% rownames_to_column("sample") %>% separate(sample, into =c("genotype", "replicate"))

mydgelist$samples <- mydgelist$samples %>%  rownames_to_column("sample") %>%
  separate(sample, into = c("Group", "Population"), sep = "_")

mydgelist_cpm <- cpm(mydgelist$counts)

#Coloured by Group

res.pca <- prcomp(t(mydgelist$counts))
#colors <- as.factor(str_split(rownames(mydgelist$samples), "_", simplify = TRUE)[,1])

#Grouped by Group
fviz_pca_ind(res.pca,
             palette = "Set1",
             col.ind = mydgelist$samples$Group,
             pointshape = 19,   # <-- THIS is the key argument
             pointsize = 2,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )

#Grouped by Population
fviz_pca_ind(res.pca,
             palette = NULL,
             col.ind = mydgelist$samples$Population,
             pointshape = 19,   # <-- THIS is the key argument
             pointsize = 2,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )

#Grouped by Group in colour and Population by shape - Not Working
fviz_pca_ind(res.pca,
             palette = NULL,
             col.ind = mydgelist$samples$Population,
             pointshape = 19,   # <-- THIS is the key argument
             pointsize = 2,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )

fviz_pca_ind(res.pca,
             palette = NULL,
             col.ind = mydgelist$samples$Group,
             habillage = mydgelist$samples$Population,
             pointsize = 2,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             ) 

mydgelist_samples$Population

?fviz_pca_ind()
#Cd49b+, Cd49b-, Cd150-, MkP, LMPP, GMP

#Coloured by Population
colors <- mydgelist_samples$Population

fviz_pca_ind(res.pca,
             palette = "Set1",
             col.ind = mydgelist_samples$Population,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             )

#Coloured by Group and Population

fviz_pca_ind(res.pca,
             palette = "Set1",
             col.ind = mydgelist_samples$Group,
             habillage = mydgelist_samples$Population,
             repel = FALSE,    # Avoid text overlapping
             title = "PCA plot of human data",
             label ="none",
             mean.point = FALSE
             ) 


#Lin; B-cells, T-cells, M-cells, NK-cells





fviz_pca_ind(
  res.pca,
  palette = "Set1",
  repel = TRUE,
  title = "PCA plot of human data"
)

fviz_pca_ind(
  res.pca,
  habillage = pseudo_3_group_population$group,  # <- factor variable
  palette = "Set1",
  repel = TRUE,
  title = "PCA plot of human data"
)

```

```{r}
p <- fviz_pca_ind(
  res.pca,
  habillage = mydgelist_samples$Group,
  palette = "Set1",
  repel = FALSE,
  label = "none",
  mean.point = FALSE
)

p$data$Population <- mydgelist_samples$Population

p +
  geom_point(aes(shape = Population)) +
  labs(color = "Group", shape = "Population")
```


